Type-Safe Modular Parsing

### 2016-11-08 paper/sections/CaseStudy.tex
Text

For the execution time comparison, we built a generator to randomly generate valid expressions according to the syntax of each calculus. Each test file corresponds with one calculus and it consists of 500 expressions randomly generated. The size of test files varies from 20KB to 100KB. After testing both implementation, we found that our version is about \huang{60\%} slower. Part of the reason is that our library requires more intermediate function calls, and the object algebra and self-reference needed to be passed as arguments every time. However, another important reason is that, because of the modular setting, we use a different alternative combinator which will try both branches and match the longest one. \huang{ref to somewhere before?} In contrast Ilya uses the ordinary alternative combinator which tries the second branch only if the first one fails. Although this difference slows our version, we want to mention that because of the memorization of Packrat parsers, we are just constant times slower, the overall complexity is still polynomial regarding to the length of the input.

>>>>

For the execution time comparison, we built a generator to randomly generate valid expressions according to the syntax of each calculus. Each test file corresponds with one calculus and it consists of 500 expressions randomly generated. The size of test files varies from 20KB to 100KB. After testing both implementation, we found that our version is about \huang{60\%} slower. Part of the reason is that our library requires more intermediate function calls, and the object algebra and self-reference needed to be passed as arguments every time. However, another important reason is that, because of the modular setting, we use a different alternative combinator which will try both branches and match the longest one. \huang{ref to somewhere before?} In contrast Ilya uses the ordinary alternative combinator which tries the second branch only if the first one fails. Although this difference slows our version, we want to mention that because of the memorization of Packrat parsers, we are just constant times slower, the big-O complexity is still the same.

### 2016-11-08 paper/sections/Overview.tex
Text

There are variants in the category of parser combinators. Almost all of them adopt top-down, recursive descent parsing, which uses backtracking to search through the possible branches. Simple backtracking parser combinators, such as the famous library Parsec in Haskell, have some drawbacks. One is that they cannot support left-recursive grammars. The common solution is to rewrite a left-recursive grammar into an equivalent one, so called left-recursion elimination. This solution requires the whole grammar, so that it cannot be applied in a modular setting because we only have parts of the grammar. Another drawback is performance.

>>>>

There are variants in the category of parser combinators. Almost all of them adopt top-down, recursive descent parsing, which uses backtracking to search through the possible branches. Simple backtracking parser combinators, such as the famous library Parsec in Haskell, have some drawbacks. One is that they cannot support left-recursive grammars. The common solution is to rewrite a left-recursive grammar into an equivalent one, so called left-recursion elimination. This solution requires information of the whole grammar, so that it cannot be applied in a modular setting because we only have parts of the grammar. Another drawback is performance. Take Parsec as an example, its choice combinator only tries the second alternative if the first fails without any token consumption. If two alternatives have a same non-empty prefix, an auxiliary function \lstinline{try} must be added to backtrack. However, if we want to extend our parsers later, we should always consider the worst case in which all alternatives share common prefixes. Then we need to add \lstinline{try} for all the branches, which results worst case exponential time complexity.

Beyond the simple backtracking one, several advanced parser combinators were studied to resolve those issues. One of them is Packrat parsers, which is our choice. The details of Packrat parser are beyond the scope of this paper. In brief, Packrat parsers use memorization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated, and it supports left-recursive grammars. These properties are very suitable for modularity. Thus we use Packrat parsers as the underlying parsing technique in our library.

It is worth mentioning that the choice of parser combinators will not affect the other parts of our library. One can choose other parser combinators like Parsec, in cases that the performance and supporting of left-recursion are not major concerns.

The second part of our library is open recursion. It is introduced to deal with the problem of recursive calls in modular parsing. We use a simple example to demonstrate the problem, which is parsing a language of natural numbers. The language is defined below.

\begin{lstlisting}[language=PlainCode]
Expr := zero | succ Expr
\end{lstlisting}

Firstly, we need to define the abstract syntax. Here we use Scala's case classes to save spaces, this can also be done using class inheritance.

\begin{lstlisting}
abstract class Expr
case class Zero extends Expr
case class Succ(v: Expr) extends Expr
\end{lstlisting}

And then we write parsing functions for every cases. Some irrelevant details in the code are omitted to outline the main structure. Suppose we have \lstinline{choice} combinator for alternatives and the input tokens are global values which can be accessed by all these functions.

\begin{lstlisting}
def parseExpr(): Expr = choice(parserZero, parseSucc)

def parserZero(): Expr = {...}

def parseSucc(): Expr = {
  parseToken('succ')
  val v = parseExpr()
  Succ(v)
}
\end{lstlisting}

It works well, but later we want to extend the language to support addition.

\begin{lstlisting}[language=PlainCode]
Expr := zero | succ Expr | add Expr Expr
\end{lstlisting}

Extending the abstract syntax is easy.

\begin{lstlisting}
case class Add(a: Expr, b: Expr) extends Expr
\end{lstlisting}
